---
title: "Evaluating Heat Adaptation Interventions"
subtitle: "A Hands-On Workshop: Case-Crossover, Difference-in-Differences, and Inverse Probability Weighting"
author: "Dr. Amruta Nori-Sarma"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output:
  bookdown::gitbook:
    split_by: chapter
    number_sections: true
    toc_depth: 2
    config:
      toc:
        collapse: section
        scroll_highlight: true
documentclass: book
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 4.5)
```

# Welcome {-}

**A note on the data before we start.** Everything in this workshop is
**simulated**. It is built to loosely resemble the scale and population
characteristics reported in the PRIME-Boston Boston Emergency Services Team
(BEST) papers (roughly 6,500 to 8,200 psychiatric ED encounters per warm
season, a population that is predominantly publicly insured or uninsured,
individually geocoded and linked to gridded ambient temperature), but the
specific counts, effect sizes, and the "true" intervention effect below are
all invented for teaching purposes. Nothing here is a re-analysis of real
BEST data, and nothing here should be read as an estimate of whether any
real intervention worked. The point of the exercise is methodological: we
know the ground truth we built into the simulation, so we can watch each
design either recover it or fail to.

**Learning objectives.** By the end of this workshop you will be able to:

1. Build a synthetic person-day dataset with realistic structure: a daily
   heat exposure series, drifting population composition, a secular
   trend, and a known intervention effect.
2. Implement a time-stratified case-crossover analysis and explain what
   it does and does not control for.
3. Implement a difference-in-differences (DiD) analysis through a
   reusable `model_generalized()` function, and build a proper
   event-study plot with year-by-year estimates relative to the
   intervention.
4. Build inverse probability weights to address compositional drift in a
   pre/post comparison, and feed them into the same `model_generalized()`
   function used for the DiD estimate.
5. Compare all three estimates against the known ground truth and discuss
   which assumptions each design leans on, tying back to the critique of
   the Ahmedabad HAP evaluation.

**Setup.** Run this once before the workshop. If you don't already have
these packages, uncomment the install line.

```{r setup}
# install.packages(c("bookdown", "dplyr", "tidyr", "tibble", "lubridate",
#                     "survival", "MASS", "timeDate", "ggplot2", "broom"))
library(lubridate)
library(survival)   # clogit() for the case-crossover
library(MASS)       # kept for the namespace-collision note below (not otherwise used)
library(timeDate)   # holiday() for the DiD models' holiday flag
library(tidyr)      # pivot_wider() / unite()
library(ggplot2)
library(broom)      # tidy() for clean model summaries
library(dplyr)      # loaded LAST on purpose, see note below

set.seed(2026)
```

**A namespace note, since it trips people up constantly.** `MASS` exports
its own `select()` function (for stepwise model selection), completely
unrelated to `dplyr::select()`. Whichever package you `library()` *last*
wins any naming collision, so if you load `MASS` after `dplyr`, every
`select(...)` call in your script silently starts calling `MASS::select()`
instead, and you'll get a confusing `unused arguments` error the moment
you try to pick columns. Loading `dplyr` last (as above) fixes it, but the
safest habit is to just write `dplyr::select()` explicitly anywhere it
matters, which is what we do below.


# Building the Synthetic BEST-like Population (10 min)

We will simulate 11 warm seasons (May 1 to September 30), 2010 to 2020,
with the intervention (our stand-in "HAP") implemented at the start of
2016. That gives us a 6-year pre-period and a 5-year post-period, similar
in spirit to the pre/post split used in the Boston resilience-plan paper
and the Ahmedabad HAP evaluation.

## A Daily Temperature Series with a Known "Extreme Heat" Threshold

We build a smooth seasonal curve, add year-to-year variability, splice in
a handful of multi-day heat wave events each year, and add daily noise.
None of this is calibrated to real Boston climatology; it is just meant to
look like a plausible warm-season temperature series with a right skew
(occasional heat waves), which is the feature that matters for the
analysis below.

```{r temperature-series}
years <- 2010:2020

build_season_dates <- function(yr) {
  seq(as.Date(paste0(yr, "-05-01")), as.Date(paste0(yr, "-09-30")), by = "day")
}

daily_df <- lapply(years, function(yr) {
  dates <- build_season_dates(yr)
  doy_in_season <- as.numeric(dates - min(dates)) + 1
  n_days <- length(dates)

  # smooth seasonal hump, peaking mid-season
  seasonal_mean <- 22 + 8 * sin(pi * doy_in_season / max(doy_in_season))

  # each year is a bit hotter or cooler than average, at random
  year_offset <- rnorm(1, mean = 0, sd = 1.2)

  # 1 to 3 heat wave events per year, each lasting 2 to 5 days, +4 to +8C
  n_waves <- sample(1:3, 1)
  wave_boost <- rep(0, n_days)
  for (w in seq_len(n_waves)) {
    wave_start <- sample(10:(n_days - 6), 1)
    wave_len <- sample(2:5, 1)
    wave_idx <- wave_start:(wave_start + wave_len - 1)
    wave_boost[wave_idx] <- wave_boost[wave_idx] + runif(1, 4, 8)
  }

  temp <- seasonal_mean + year_offset + wave_boost + rnorm(n_days, 0, 1.5)

  data.frame(date = dates, year = yr, month = month(dates), temp = temp)
}) %>% bind_rows()

# define "extreme heat" as the 90th percentile of the full 2010-2020 series,
# the same logic used in the PRIME-Boston papers
extreme_threshold <- quantile(daily_df$temp, 0.90)
daily_df <- daily_df %>%
  mutate(
    extreme = as.integer(temp >= extreme_threshold),
    post    = as.integer(year >= 2016)
  )

cat("Extreme heat threshold (90th percentile):", round(extreme_threshold, 1), "C\n")

ggplot(daily_df, aes(x = date, y = temp, color = factor(extreme))) +
  geom_line(color = "grey60") +
  geom_point(size = 0.6) +
  scale_color_manual(values = c("0" = "grey60", "1" = "firebrick"),
                      labels = c("Below threshold", "Extreme heat day"),
                      name = NULL) +
  geom_vline(xintercept = as.Date("2016-01-01"), linetype = "dashed") +
  labs(title = "Simulated daily warm-season temperature, 2010-2020",
       subtitle = "Dashed line marks the intervention year (2016)",
       x = NULL, y = "Temperature (C)") +
  theme_minimal()
```

## Simulating Individual Encounters, with a Known Ground Truth Baked In

Now we generate the person-level encounters. Four things are built into
the data-generating process on purpose, because each one is what makes the
three analytic methods behave differently:

- **A secular upward trend** in daily encounter volume, unrelated to heat,
  of about 2.5% per year. This is the kind of trend a naive pre/post
  comparison cannot distinguish from an intervention effect, and exactly
  the kind of confounding a DiD design is built to net out.
- **A real heat effect**: encounter counts rise with temperature above a
  25C threshold.
- **A true intervention effect**: after 2016, the heat-slope is
  attenuated by roughly a third. This is the effect we're hoping our
  analyses can recover.
- **Drifting population composition**: the share of patients who are
  Medicaid-insured or uninsured, and the share with a chronic serious
  mental illness diagnosis, both rise slowly over the study period. This
  mirrors the exact concern Hess et al. (2018) raised about Ahmedabad
  (demographic shifts between the pre- and post-periods) but did not have
  the data to address. It's what motivates the IPW step later.

```{r simulate-encounters}
# day-level expected count: baseline * trend * heat effect * (attenuated after 2016)
daily_df <- daily_df %>%
  mutate(
    year_c      = year - 2010,
    trend_term  = 0.025 * year_c,
    heat_term   = 0.050 * pmax(temp - 25, 0),
    heat_x_post = -0.018 * post * pmax(temp - 25, 0),   # TRUE effect: ~36% slope reduction
    log_lambda  = log(45) + trend_term + heat_term + heat_x_post,
    lambda      = exp(log_lambda),
    n_encounters = rpois(n(), lambda)
  )

# expand each day into individual encounter-level rows
make_day_encounters <- function(row) {
  n <- row$n_encounters
  if (n == 0) return(NULL)

  yr <- row$year
  # population composition drifts over the study period
  p_medicaid_uninsured <- 0.70 + 0.015 * (yr - 2010)   # rising, capped below
  p_medicaid_uninsured <- min(p_medicaid_uninsured, 0.92)
  insurance_probs <- c(
    Medicaid    = p_medicaid_uninsured * 0.65,
    Uninsured   = p_medicaid_uninsured * 0.35,
    Medicare    = (1 - p_medicaid_uninsured) * 0.55,
    Commercial  = (1 - p_medicaid_uninsured) * 0.45
  )

  p_chronic_mi <- min(0.28 + 0.006 * (yr - 2010), 0.45)   # rising chronic illness share
  svi_mean     <- min(0.45 + 0.01  * (yr - 2010), 0.70)   # rising social vulnerability

  data.frame(
    date         = rep(row$date, n),
    year         = yr,
    month        = row$month,
    temp         = row$temp + rnorm(n, 0, 0.5),   # small within-city spatial noise
    extreme      = row$extreme,
    post         = row$post,
    age          = pmin(pmax(round(rnorm(n, 38, 13)), 18), 90),
    sex          = sample(c("Male", "Female"), n, replace = TRUE, prob = c(0.53, 0.47)),
    insurance    = sample(names(insurance_probs), n, replace = TRUE, prob = insurance_probs),
    chronic_mi   = rbinom(n, 1, p_chronic_mi),
    svi          = pmin(pmax(rnorm(n, svi_mean, 0.12), 0), 1)
  )
}

encounter_list <- lapply(seq_len(nrow(daily_df)), function(i) make_day_encounters(daily_df[i, ]))
encounters <- bind_rows(encounter_list)
encounters$id <- seq_len(nrow(encounters))

cat("Total simulated encounters:", nrow(encounters), "\n")
cat("Average per warm season:", round(nrow(encounters) / length(years)), "\n")
cat("(Compare to the real BEST scale reported in the Boston DiD papers:",
    "roughly 6,500-8,200 encounters per warm season)\n")
```

## Sanity Checks

Always look at your simulated data before you trust an analysis built on
top of it.

```{r sanity-checks}
# annual encounter volume: should show the ~2.5%/year upward trend
encounters %>%
  count(year) %>%
  ggplot(aes(x = year, y = n)) +
  geom_col(fill = "#A51C30") +
  geom_vline(xintercept = 2015.5, linetype = "dashed") +
  labs(title = "Simulated annual encounter volume",
       subtitle = "Dashed line marks the intervention year",
       x = NULL, y = "Encounters") +
  theme_minimal()

# insurance mix drifting over time, exactly the compositional shift
# Hess et al. flagged as an unmeasured concern for Ahmedabad
encounters %>%
  count(year, insurance) %>%
  group_by(year) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(x = year, y = pct, fill = insurance)) +
  geom_area() +
  labs(title = "Simulated insurance mix over time",
       subtitle = "Medicaid/uninsured share rises across the study period",
       x = NULL, y = "Share of encounters") +
  theme_minimal()
```


# Method 1: Time-Stratified Case-Crossover

Each case (an encounter) is matched to a handful of control days: same
person's location, same calendar month and year, same day of week as the
case day, excluding the case day itself. Because each person is compared
only to their own control days, this design controls for anything that is
constant for that person and stratum by design (age, insurance, chronic
illness, neighborhood). It does **not** control for a citywide secular
trend across the full study period, because that trend operates just as
much on the control days as the case day within a given month.

To keep runtimes reasonable in a workshop setting, we subsample cases
rather than using all ~75,000+ simulated encounters. Feel free to raise
`n_cases_per_period` and rerun at home.

```{r case-crossover-build}
n_cases_per_period <- 800

pre_cases  <- encounters %>% filter(post == 0) %>% slice_sample(n = n_cases_per_period)
post_cases <- encounters %>% filter(post == 1) %>% slice_sample(n = n_cases_per_period)
sampled_cases <- bind_rows(pre_cases, post_cases)

# lookup table of city-wide daily exposure, used for both case and control days
temp_lookup <- daily_df %>% dplyr::select(date, temp, extreme)

get_control_days <- function(case_date, n_controls = 4) {
  same_month <- format(case_date, "%Y-%m")
  candidates <- temp_lookup %>%
    filter(format(date, "%Y-%m") == same_month,
           wday(date) == wday(case_date),
           date != case_date)
  if (nrow(candidates) > n_controls) {
    candidates <- candidates %>% slice_sample(n = n_controls)
  }
  candidates
}

build_cc_dataset <- function(cases_df) {
  out <- vector("list", nrow(cases_df))
  for (i in seq_len(nrow(cases_df))) {
    case_row <- cases_df[i, ]
    controls <- get_control_days(case_row$date)
    if (nrow(controls) == 0) next

    case_line <- data.frame(
      strat_id = case_row$id, case = 1,
      temp = case_row$temp, extreme = case_row$extreme
    )
    control_lines <- data.frame(
      strat_id = case_row$id, case = 0,
      temp = controls$temp, extreme = controls$extreme
    )
    out[[i]] <- bind_rows(case_line, control_lines)
  }
  bind_rows(out)
}

cc_pre  <- build_cc_dataset(pre_cases)
cc_post <- build_cc_dataset(post_cases)

cat("Pre-period case-crossover rows:", nrow(cc_pre), "\n")
cat("Post-period case-crossover rows:", nrow(cc_post), "\n")
```

```{r case-crossover-fit}
cc_model_pre  <- clogit(case ~ extreme + strata(strat_id), data = cc_pre)
cc_model_post <- clogit(case ~ extreme + strata(strat_id), data = cc_post)

cc_summary <- bind_rows(
  tidy(cc_model_pre,  exponentiate = TRUE, conf.int = TRUE) %>% mutate(period = "Pre (2010-2015)"),
  tidy(cc_model_post, exponentiate = TRUE, conf.int = TRUE) %>% mutate(period = "Post (2016-2020)")
) %>%
  filter(term == "extreme") %>%
  select(period, OR = estimate, conf.low, conf.high, p.value)

print(cc_summary)
```

**What this does and does not tell us.** Comparing the pre-period OR to
the post-period OR gives an informal before/after signal, similar in
spirit to comparing Ahmedabad's pre-HAP and post-HAP relative risks. But
notice that we have not formally tested whether the two ORs differ from
each other, and the case-crossover design by itself has no mechanism for
netting out a citywide secular trend that shifted between the two multi-
year windows. That's exactly the gap the next method is built to close.

# Method 2: Difference-in-Differences

This section follows a three-step DiD pipeline: a descriptive "Table 1"
style pre/post comparison with a naive percentage-based DiD term, an
event-study plot with a proper event-time regression, and then the
formal Poisson regression, run through a reusable
`model_generalized()` function.

```{r did-build}
# a few extra day-level covariates the models below expect: day of week,
# week number, and a simple summer holiday flag
holiday_years <- 2010:2020
holidays <- c(
  as.Date(timeDate::holiday(holiday_years, "USMemorialDay")),
  as.Date(timeDate::holiday(holiday_years, "USIndependenceDay")),
  as.Date(timeDate::holiday(holiday_years, "USLaborDay"))
)

daily_counts <- encounters %>%
  count(date, name = "n_encounters_actual") %>%
  right_join(daily_df, by = "date") %>%
  mutate(
    n_encounters_actual = ifelse(is.na(n_encounters_actual), 0, n_encounters_actual),
    dow   = wday(date),
    week  = week(date),
    holid = as.integer(date %in% holidays)
  )
```

## 2.1 Descriptive Table 1, with a naive DiD term

Compute each group's share of its own period's total, take pre-to-post
differences within the extreme and non-extreme groups separately, then
difference those two differences, a quick sanity check to run before
turning to any regression model.

```{r did-table1}
pre_total  <- sum(daily_counts$n_encounters_actual[daily_counts$post == 0])
post_total <- sum(daily_counts$n_encounters_actual[daily_counts$post == 1])

did_table1 <- daily_counts %>%
  mutate(extreme_chr = ifelse(extreme == 1, ">= 90th pctile", "< 90th pctile"),
         post_chr    = ifelse(post == 1, "Post", "Pre")) %>%
  group_by(extreme_chr, post_chr) %>%
  summarise(visits = sum(n_encounters_actual), .groups = "drop") %>%
  tidyr::unite("group", extreme_chr, post_chr, sep = " ") %>%
  tidyr::pivot_wider(names_from = group, values_from = visits)

did_table1 <- did_table1 %>%
  mutate(
    `< 90th pctile Pre (%)`   = round(100 * `< 90th pctile Pre`   / pre_total, 1),
    `< 90th pctile Post (%)`  = round(100 * `< 90th pctile Post`  / post_total, 1),
    `>= 90th pctile Pre (%)`  = round(100 * `>= 90th pctile Pre`  / pre_total, 1),
    `>= 90th pctile Post (%)` = round(100 * `>= 90th pctile Post` / post_total, 1),
    `< 90th pctile Diff`  = `< 90th pctile Post (%)`  - `< 90th pctile Pre (%)`,
    `>= 90th pctile Diff` = `>= 90th pctile Post (%)` - `>= 90th pctile Pre (%)`,
    `DiD Term` = `< 90th pctile Diff` - `>= 90th pctile Diff`
  )

print(did_table1, width = Inf)
```

## 2.2 Event-study plot

A proper event-study design: a Poisson regression with the extreme-heat
indicator interacted with a full set of year-relative-to-intervention
dummies (event time), so each year gets its own estimated heat effect
relative to the year immediately before the intervention (event time
-1, the omitted reference category), rather than just comparing raw
annual means.

```{r did-event-study}
event_study_plot <- function(dataframe, outcome_variable, threshold_temp,
                              intervention_year, start_year, end_year) {

  data <- dataframe %>%
    filter(year %in% start_year:end_year) %>%
    mutate(event_time = year - intervention_year)

  list_event_times <- as.character(seq(min(data$event_time), max(data$event_time)))
  list_event_times <- c("-1", list_event_times[list_event_times != "-1"])

  data <- data %>%
    mutate(event_time_factor = factor(event_time, levels = list_event_times))

  data$outcome_variable <- data[[outcome_variable]]
  data$threshold_temp   <- data[[threshold_temp]]

  event_model <- glm(
    outcome_variable ~ threshold_temp +
      (threshold_temp:event_time_factor) +
      event_time_factor +
      holid + year + dow + month + week,
    data = data,
    family = poisson
  )

  coefficients <- coef(event_model)
  interaction_coefficient <- coefficients[grep("threshold_temp:event_time_factor", names(coefficients))]

  conf_intervals <- confint(event_model)
  interaction_confint <- conf_intervals[grep("threshold_temp:event_time_factor", rownames(conf_intervals)), ]

  plot_data <- data.frame(cbind(interaction_coefficient, interaction_confint)) %>%
    tibble::rownames_to_column(var = "EventTime")
  plot_data$EventTime <- as.numeric(sub("^threshold_temp:event_time_factor", "", plot_data$EventTime))
  plot_data$interaction_coefficient <- exp(plot_data$interaction_coefficient)
  plot_data$lowCI <- exp(plot_data$X2.5..)
  plot_data$uppCI <- exp(plot_data$X97.5..)

  new_row <- data.frame(EventTime = -1, interaction_coefficient = 1, lowCI = 1, uppCI = 1)
  plot_data <- plot_data %>% dplyr::select(-c(X2.5.., X97.5..))
  plot_data2 <- rbind(plot_data, new_row)

  ggplot(plot_data2, aes(EventTime, interaction_coefficient)) +
    geom_line() +
    geom_point() +
    geom_errorbar(aes(ymin = lowCI, ymax = uppCI)) +
    geom_vline(xintercept = 0, color = "#A51C30", linetype = "dashed", linewidth = 1) +
    geom_hline(yintercept = 1, color = "grey") +
    xlab("Time to intervention (years)") +
    ylab("Estimated heat effect, relative to year -1") +
    theme_minimal() +
    scale_x_continuous(
      breaks = min(plot_data2$EventTime):max(plot_data2$EventTime),
      labels = as.character(min(plot_data2$EventTime):max(plot_data2$EventTime))
    )
}

event_study_plot(
  dataframe = daily_counts, outcome_variable = "n_encounters_actual",
  threshold_temp = "extreme", intervention_year = 2016,
  start_year = 2010, end_year = 2020
)
```

**Reading the plot.** Each point is a year's estimated multiplicative
effect of an extreme heat day on encounter counts, relative to the year
right before the intervention. If the intervention worked, you'd expect
the post-intervention points (event time >= 0) to sit visibly lower than
the pre-intervention points. Flat, overlapping confidence intervals
across the dashed line, on the other hand, look a lot like the pattern
Hess et al. found when they tested whether Ahmedabad's effect held up
without the anomalous 2010 heat wave in the baseline.

## 2.3 Poisson regression via `model_generalized()`

A reusable function for the main DiD estimate: a Poisson regression of
the outcome on exposure, intervention period, their interaction, and day
of week, week, month, year, and a holiday flag, returning the interaction
term as an incidence rate ratio with its 95% profile-likelihood
confidence interval.

```{r did-model-generalized-fn}
model_generalized <- function(dataframe, outcome_count, exposure, intervention, weights = NULL) {
  if (is.null(weights)) {
    model <- glm(
      outcome_count ~ exposure + intervention + (exposure:intervention) +
        year + holid + month + week + dow,
      data = dataframe, family = poisson
    )
  } else {
    model <- glm(
      outcome_count ~ exposure + intervention + (exposure:intervention) +
        year + holid + month + week + dow,
      data = dataframe, family = quasipoisson, weights = weights
    )
  }

  coefficients <- coef(model)
  interaction_coefficient <- exp(coefficients[grep("exposure:intervention", names(coefficients))])

  conf_intervals <- exp(confint(model))
  interaction_confint <- conf_intervals[grep("exposure:intervention", rownames(conf_intervals)), ]

  paste0(round(interaction_coefficient, 2), " (",
         round(interaction_confint[1], 2), ", ",
         round(interaction_confint[2], 2), ")")
}
```

**Overall model:**

```{r did-overall}
did_overall_irr <- model_generalized(
  dataframe    = daily_counts,
  outcome_count = daily_counts$n_encounters_actual,
  exposure     = daily_counts$extreme,
  intervention = daily_counts$post
)
did_overall_irr
```

**Stratified calls**, applied to a couple of subgroups the way an actual
stratified analysis would: insurance type and chronic mental illness
status, since those are the two characteristics with drifting composition
built into the simulation.

```{r did-stratified}
daily_by_group <- function(filter_expr) {
  encounters %>%
    filter({{ filter_expr }}) %>%
    count(date, name = "n_encounters_actual") %>%
    right_join(daily_df, by = "date") %>%
    mutate(
      n_encounters_actual = ifelse(is.na(n_encounters_actual), 0, n_encounters_actual),
      dow   = wday(date),
      week  = week(date),
      holid = as.integer(date %in% holidays)
    )
}

medicaid_uninsured  <- daily_by_group(insurance %in% c("Medicaid", "Uninsured"))
commercial_medicare <- daily_by_group(insurance %in% c("Commercial", "Medicare"))
chronic_mi_group    <- daily_by_group(chronic_mi == 1)

data.frame(
  subgroup = c("Medicaid/Uninsured", "Commercial/Medicare", "Chronic mental illness"),
  IRR = c(
    model_generalized(medicaid_uninsured, medicaid_uninsured$n_encounters_actual,
                       medicaid_uninsured$extreme, medicaid_uninsured$post),
    model_generalized(commercial_medicare, commercial_medicare$n_encounters_actual,
                       commercial_medicare$extreme, commercial_medicare$post),
    model_generalized(chronic_mi_group, chronic_mi_group$n_encounters_actual,
                       chronic_mi_group$extreme, chronic_mi_group$post)
  )
)
```

**Reading the output.** The interaction-term IRR is the DiD estimate: a
value below 1 means the extra risk on an extreme heat day was smaller
after the intervention than before it, controlling for day of week,
week, month, year, and holidays. Because we built a true attenuation of
about 36% into the simulation, a well-specified DiD model should land
somewhere in that neighborhood for the overall estimate.

# Method 3: Inverse Probability Weighting for Pre/Post Comparison

The DiD model above still assumes the population being served didn't
change in ways that matter. But we built in exactly the kind of
compositional drift Hess et al. flagged as an unmeasured concern for
Ahmedabad: rising Medicaid/uninsured share, rising chronic mental illness
prevalence, rising social vulnerability. If those shifts are themselves
associated with heat sensitivity, they could bias the DiD estimate in
either direction.

The IPW approach: model the probability that a given encounter falls in
the post-period as a function of its measured characteristics, then
reweight each day's contribution to the model so that the weighted
population looks similar across the pre- and post-periods on those
measured characteristics. This does not fix confounding by *unmeasured*
factors (air pollution, healthcare access, publicity effects) any more
than the original Ahmedabad study could, but it does directly address the
one confounding pathway we actually simulated and can observe.

```{r ipw-build}
ps_model <- glm(
  post ~ age + sex + insurance + chronic_mi + svi,
  data = encounters, family = binomial
)

encounters$p_post <- predict(ps_model, type = "response")
marginal_p_post <- mean(encounters$post)

encounters <- encounters %>%
  mutate(
    sw = ifelse(post == 1, marginal_p_post / p_post, (1 - marginal_p_post) / (1 - p_post))
  )

# trim extreme weights, standard practice to avoid a small number of
# observations dominating the model
sw_caps <- quantile(encounters$sw, c(0.01, 0.99))
encounters$sw_trimmed <- pmin(pmax(encounters$sw, sw_caps[1]), sw_caps[2])

summary(encounters$sw_trimmed)
```

```{r ipw-balance-check}
# quick covariate balance check: unweighted vs weighted mean SVI and
# chronic_mi by period. If weighting is doing its job, the weighted
# pre/post gap should shrink relative to the unweighted gap.
balance_check <- encounters %>%
  group_by(post) %>%
  summarise(
    unweighted_svi        = mean(svi),
    weighted_svi          = weighted.mean(svi, sw_trimmed),
    unweighted_chronic_mi = mean(chronic_mi),
    weighted_chronic_mi   = weighted.mean(chronic_mi, sw_trimmed)
  )

print(balance_check)
```

```{r ipw-fit}
# aggregate the per-encounter weights up to a day-level average weight,
# then refeed the SAME model_generalized() function used for the DiD
# section, now with weights supplied
day_weights <- encounters %>%
  group_by(date) %>%
  summarise(day_avg_sw = mean(sw_trimmed))

daily_counts_ipw <- daily_counts %>% left_join(day_weights, by = "date")
daily_counts_ipw$day_avg_sw[is.na(daily_counts_ipw$day_avg_sw)] <- 1  # no encounters that day

ipw_irr <- model_generalized(
  dataframe    = daily_counts_ipw,
  outcome_count = daily_counts_ipw$n_encounters_actual,
  exposure     = daily_counts_ipw$extreme,
  intervention = daily_counts_ipw$post,
  weights      = daily_counts_ipw$day_avg_sw
)
ipw_irr
```

**What to look for.** Compare this IRR to the unweighted DiD estimate
from Section 2.3. If they're close, that's evidence the compositional
drift we simulated wasn't actually biasing the unweighted estimate much.
If they diverge, that tells you the population shift was doing some of
the work you might otherwise have attributed to the intervention itself,
exactly the ambiguity Hess et al. could not rule out for Ahmedabad
because they had no individual-level data to weight on.

# Bringing It Together

```{r comparison-table}
true_effect_note <- "True simulated effect: ~36% reduction in heat-slope after 2016"

comparison <- data.frame(
  Method = c("Case-crossover (informal pre vs post OR comparison)",
             "Difference-in-differences (unweighted, model_generalized())",
             "Difference-in-differences (IPW-weighted, model_generalized())"),
  Estimate = c(
    paste0(round(cc_summary$OR[cc_summary$period == "Pre (2010-2015)"], 2), " -> ",
           round(cc_summary$OR[cc_summary$period == "Post (2016-2020)"], 2)),
    did_overall_irr,
    ipw_irr
  ),
  What_it_controls_for = c(
    "Time-invariant confounders per person and stratum, not the citywide secular trend",
    "Shared secular trend and season, not compositional drift",
    "Shared secular trend, season, and measured compositional drift"
  )
)

print(comparison, row.names = FALSE)
cat("\n", true_effect_note, "\n")
```


**Discussion questions for the group:**

1. Which of these three estimates would you trust most, and why? Is your
   answer different if you imagine this were real data versus knowing, as
   we do here, the actual ground truth?
2. The case-crossover comparison is the closest analogue in this workshop
   to what a simple pre/post design (like the core of the Ahmedabad HAP
   evaluation) can offer. What would you need to add to make it a genuine
   causal comparison?
3. The IPW step only adjusts for compositional drift we measured and
   simulated on purpose. What are two or three plausible *unmeasured*
   confounders, in either Boston or Ahmedabad, that no amount of
   reweighting on the variables we have could fix?
4. If you had to design a real evaluation of a new heat-adaptation program
   with only 2 years of pre-data and 2 years of post-data, which of these
   three designs would you prioritize building the data infrastructure
   for, and why?


# Session Info {-}

```{r session-info}
sessionInfo()
```
